---
title: "In class assignment 1"
author: "Dr. Nabila"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
# Built-in data only (no downloads, no API keys)
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```

# 0) What is a time series in R?

A **time series** is data recorded over time.  
In R, a `ts` object stores:
- the data
- start/end time
- frequency (e.g. 12 = monthly)

```{r}
ap <- AirPassengers
class(ap)
frequency(ap)
start(ap); end(ap)
```

---

# 1) AirPassengers

```{r}
plot(ap,
     main = "AirPassengers: Trend + Seasonality",
     ylab = "Passengers (thousands)",
     xlab = "Year")
```

### 🟦 PAUSE & DISCUSS (2–3 minutes)
- Do you see an **overall direction**?  
** We see an upward trend. 

- Do you see a **repeating pattern**?  
** Yes we do see a seasonal pattern.


- Why might airline travel repeat every year?
** Because towards the summer, there tends to be a lot of tourists traveling, while colder seasons people travel less. 

---

## 1A) Seasonality made obvious

```{r}
monthplot(ap,
          main = "AirPassengers: Month Plot (Seasonality)",
          ylab = "Passengers (thousands)")
```

### 🟦 PAUSE & DISCUSS
- Which months are consistently high? Low?
** First the Jan, Feb, and Mar, and also Oct, Nov, and Dec are low. At the same time, 
from April, it gets higher until September. 


- What real-world behavior could explain this?
** It could be the behavior of tourists becuase it starts getting higher from the begening of summer and is high up until the end fall. 

---

## 1B) Decomposition

```{r}
ap_decomp <- decompose(ap, type = "multiplicative")
plot(ap_decomp)
```

### 🟦 PAUSE & DISCUSS
- What does the **trend** component capture?
**  Increase in the number of passengers compared to the previous year. 


- What does the **seasonal** component assume?
** Shows that on particular time of the year, the passenger number peaks, and decreases.


- What might the **remainder** represent?
** It can represents the randomness in the increase/decrease of the passengers that doesn't fit any of the category above. 


---

# 2) Trend / Level Shifts (Nile)

```{r}
nile <- Nile
plot(nile,
     main = "Nile River Flow: Trend / Level Shifts",
     ylab = "Flow",
     xlab = "Year")
abline(h = mean(nile), lty = 2)
```

### 🟦 PAUSE & DISCUSS
- Does this series look seasonal?
** It does not look seasonal. 


- Do you notice **level changes**?
** Yes it does change in its level. 


- Why might long-run averages shift in environmental data?
** It can be due to climate change or climate pattern of a specific region. 

---

# 3) Cyclical behavior (UKgas)

```{r}
ukg <- UKgas
frequency(ukg)

plot(ukg,
     main = "UKgas: Cyclical Movements (Quarterly)",
     ylab = "Gas consumption",
     xlab = "Year")
```

### 🟦 PAUSE & DISCUSS
- Are the ups and downs evenly spaced?
** Yes, they are evenly spaced by 10 years. 

- How is this different from seasonality?
** Because the lower and upper parts' symmetry differ from each other so we have two different seasonality. 

- What might link this series to the business cycle?
** Maybe gas got cheaper. 


---

## 3A) Differencing to highlight cycles

```{r}
ukg_diff <- diff(ukg)
plot(ukg_diff,
     main = "UKgas: First Differences",
     ylab = "Change in consumption",
     xlab = "Year")
abline(h = 0, lty = 2)
```

### 🟦 PAUSE & DISCUSS
- What did differencing remove?
- Why do economists often difference data?

---

## 3B) Another cyclical example (monthly deaths)

```{r}
lde <- ldeaths
plot(lde,
     main = "ldeaths: Cyclical / Seasonal Pattern",
     ylab = "Deaths",
     xlab = "Year")
```

### 🟦 PAUSE & DISCUSS
- Is this pattern purely seasonal?
- What external factors could drive cycles here?

---

# 4) Heteroscedasticity / Volatility (EuStockMarkets)

```{r}
esm <- EuStockMarkets
dax <- esm[, "DAX"]

plot(dax,
     main = "DAX Index Level",
     ylab = "Index level",
     xlab = "Time")
```

### 🟦 PAUSE & DISCUSS
- Does variance look constant here?
- Why might prices hide volatility patterns?

---

## 4A) Returns show changing variance

```{r}
dax_ret <- diff(log(dax))
plot(dax_ret,
     main = "DAX Log Returns: Volatility Clustering",
     ylab = "Return",
     xlab = "Time")
abline(h = 0, lty = 2)
```

### 🟦 PAUSE & DISCUSS
- Where do you see calm periods?
- Where do you see turbulent periods?
- Why is this a problem for simple OLS models?


---

# 5) Cross-section vs Time Series

```{r}
set.seed(1)
x <- rnorm(200)
y <- rnorm(200)
plot(x, y,
     main = "Cross-section: Order Does NOT Matter",
     xlab = "X",
     ylab = "Y")
```

---

## 5A) Shuffling destroys time-series meaning

```{r}
ap_shuffled <- ap
ap_shuffled[] <- sample(ap)

plot(ap_shuffled,
     main = "AirPassengers SHUFFLED",
     ylab = "Passengers",
     xlab = "Year")
```

### 🟦 PAUSE & DISCUSS
- Same numbers, different order — what changed?
- Why is time ordering information?

---

# 6) Student playground

```{r}
plot(nottem, main="Monthly temperatures (nottem)")
plot(co2, main="Atmospheric CO2")
plot(sunspot.year, main="Sunspots (annual)")
```

### 🟦 PAUSE & DISCUSS
- Which series is seasonal?
- Which is trending?
- Which might be hardest to forecast?

---

# 7) Big takeaways 

- Time ordering matters  
- Different series have different structures  
- Plots come **before** models  
- The model choice depends on what you *see*
